\chapter{Introduction}

\section{Context}

Mathematics education has an important role in students' lives. Its primary aim is to develop fluency in mathematical fundamentals, reasoning, and problem-solving skills. It also includes applications of mathematical concepts across science and related fields~\parencite{Math_Curriculum_England}. Government reports indicate that the average class size is about 26.2 pupils~\parencite{Class_Size_England}, with an average pupil-teacher ratio of 18.0~\parencite{Pupil_Teacher_Ratio_England}. These ratios can limit the attention each pupil receives.

Learning can be divided into procedural fluency and conceptual understanding. Procedural fluency refers to the ability to carry out mathematical procedures accurately, efficiently, and flexibly. Conceptual understanding means grasping mathematical concepts in real-world or applied contexts. Early maths education in the UK stresses procedural fluency. Pupils do mental arithmetic in Key Stage 1. Calculators are introduced only at the end of Key Stage 2.

\section{Problem}

This section identifies the central problem: static practices in arithmetic education fail to adapt to individual student needs, resulting in inefficient learning and missed opportunities for targeted intervention.

\subsection{Limitations of Static Pedagogical Models}

Teaching early arithmetic, such as addition, subtraction, multiplication, and division, requires a lot of practice. The large volume of practice needed for effective procedural fluency leads to a one-size-fits-all solution: large lists of exercises. This static approach is suboptimal. A pupil may be given a previously mastered exercise, wasting their time. They may also encounter problems they cannot solve. The ideal balance of exercise difficulty is known as the Zone of Proximal Development.

The efficiency of practice may be increased through automation. Automatically adjusting difficulty levels requires an algorithm that estimates a student's mastery over time. These algorithms, which estimate a learner's knowledge state and predict future performance based on prior responses, are known as knowledge tracing models.

\subsection{Data Sensitivity}

To address this, all training and comparison data will be synthetic student data rather than collected from children.

\section{Aims and Objectives}

This dissertation aims to assess whether LLM-generated synthetic student data reliably supports the evaluation and comparison of knowledge tracing models in a smart tutor app for basic maths operations.

The research questions addressed are:

\begin{enumerate}
    \item To what extent can knowledge tracing algorithms be effectively evaluated and compared using synthetic student data generated by large language models?
    \item Which knowledge tracing model demonstrates the strongest predictive performance when both trained and tested on LLM-generated synthetic student data within the smart tutor app?
    \item To what degree do evaluation results on synthetic student data align with results on real educational data, such as ASSISTments and real app users?
\end{enumerate}

To achieve the overall aim and answer these questions, the dissertation pursues the following objectives.

The first objective is to design and implement a smart tutor app that estimates a user's mastery using knowledge tracing models. The app should train these models to predict the probability of correct answers based on past interactions and current simple maths questions, such as addition, subtraction, multiplication, or division. It should support multiple knowledge tracing algorithms via the pyKT library~\parencite{pyktPaper}. The database will store both correctness and response time for each interaction. This app will serve as the experimental environment for evaluating synthetic, ASSISTments, and real user datasets.

The second objective is to develop an LLM-based student-generation pipeline. This should generate many fully synthesised interaction sequences across models such as Qwen, Mistral, Llama, and DeepSeek. Each interaction will be stored in the database, including the correctness of the answer and the response time. Skill states are managed outside the large language model. The current skill state and response history in the prompt ensure the LLM mimics a simulated learner rather than an unconstrained generator. The script should generate at least 100 student interaction sequences per model. Each should have a random number of answered questions between 3 and 100.

The third objective is to train, evaluate, and compare multiple knowledge tracing models. There is a wide range of knowledge tracing models and LLMs. This dissertation focuses on a carefully selected, diverse subset representing the broader landscape. The comparison spans LLM-generated synthetic student datasets, the publicly available educational dataset ASSISTments2009~\autocite{assist2009}, and real user data from the smart tutor app. The compared models include Attentive Knowledge Tracing~\autocite{akt}, Deep Knowledge Tracing~\autocite{dkt}, Diagnostic Transformer~\autocite{dtransformer}, Self-Attentive Knowledge Tracing~\autocite{sakt}, and Simple Knowledge Tracing~\autocite{simpleKT}. The evaluation metrics are the area under the ROC curve (AUC) and accuracy. These assess predictive performance and relative rankings across datasets and models.

Finally, the dissertation analyses the extent to which evaluations based on LLM-generated synthetic student data align with those based on real educational data. The focus is on agreement in model rankings and on systematic performance gaps between synthetic student data and real datasets. This analysis directly addresses whether synthetic student data can act as a proxy for real data when benchmarking knowledge tracing models.

\subsection{Key Contributions}
This dissertation makes two main contributions. The first is a user synthesis system that generates educational interaction data for training and evaluating knowledge tracing algorithms. The second is a smart tutor framework that can be reused and extended in future research on intelligent tutoring systems.

The user synthesis system uses large language models (LLMs) to interpret an internal student state and generate synthetic users. The internal state is represented as a per-skill Hidden Markov Model, inspired by Bayesian Knowledge Tracing, with four interpretable parameters: current mastery, learning rate, guess rate, and slip rate. This state, together with a response history, is passed to an LLM to generate answers, and the internal state is then updated using Bayes' rule.

The smart tutor is an open, end-to-end framework that can interact with both synthetic and real users. It includes a web frontend but is designed primarily as a REST API, so the user interface can be replaced by any client that can communicate with the backend.

\subsection{Changes Since The Proposal}

Compared to the original proposal, this dissertation places greater emphasis on the generation and use of fully synthetic student data. The proposal initially intended to apply the user synthesis methodology described in

``Analysis of Knowledge Tracing Performance on Synthesised Student Data''~\autocite{pagonis2024studentsynthesis}. However, that approach primarily focuses on augmenting existing datasets. The requirements of this dissertation necessitated synthesising student data from scratch to drive the smart tutor. As a result, the objective of adapting that specific augmentation methodology was omitted. Although it could have been used to expand data obtained from real user testing, time constraints prevented its implementation.

The proposal also suggested using the EduKTM library. This was dropped in favour of pyKT, which has better community support and maintenance. While EduKTM has 181 citations and pyKT has 121, pyKT's GitHub has 395 stars. EduKTM has 254. EduKTM has not been updated in about three years. However, pyKT is still active. All data is current as of 30th April 2026. This change ensures the tutor depends on a current, well-supported KT implementation.

Other changes were necessary due to time constraints. Implementing response time as a feature in knowledge tracing models and examining its impact on predictive performance were excluded. However, response time is still recorded and generated by the user synthesis script. It is also stored in the database, although not included as a primary input to the models. No research question focuses specifically on response time. The dissertation instead emphasises correctness-based interaction data for model training and evaluation.

In summary, compared to the proposal, the dissertation places greater emphasis on generating synthetic student data to evaluate and compare knowledge tracing algorithms. It supports these efforts with a more recent KT library, rather than focusing on response-time-augmented models or data augmentation.
\section{Outline}

\textbf{Introduction:}
\textit{Introduces the context and problem of the dissertation, states the aims and research questions, and summarises the structure of the document.}

\textbf{Background research:}
\textit{Reviews prior work on the dissertation topic. Introduces knowledge tracing and selected knowledge tracing algorithms, presents the pyKT library used to implement the models, and briefly discusses large language models.}

\textbf{Methodology:}
\textit{Describes the architecture of the smart tutor system and the design of the user data synthesis pipeline. Justifies the chosen approaches and explains the setup for comparing models and datasets.}

\textbf{Results:}
\textit{Presents the implemented smart tutor system and student synthesis pipeline, the metadata and metrics of the produced datasets, and the evaluation results for the knowledge tracing models. Includes basic testing of the smart tutor system.}

\textbf{Conclusions \& Future Work:}
\textit{Summarises the main findings in relation to the research questions, discusses limitations of the system and methodology, and proposes improvements and directions for future research.}