\chapter{Limitations \& Future Work}

This chapter discusses the challenges of this dissertation, the field of knowledge tracing, and how other research can expand on it. Future work could focus on developing new data-collection methods suitable for vulnerable populations and on exploring adaptive knowledge-tracing models that accommodate ethical constraints.

\section{Data collection}

One ongoing challenge in knowledge tracing is data collection. This challenge is made worse when the problem targets children. There are many ethical considerations when working with groups of people who are unable to consent. Testing knowledge tracing algorithms is possible using existing datasets, but this limits research to developing new models. For research like this dissertation, where the question is tailored to a topic and would require specific questions to be answered, it is made difficult.

The dissertation is limited to comparing synthetic users to existing educational data, which is anonymous and targets different fields of education and anonymous adults with high prior mastery. This makes it impossible to answer specific questions about how well the knowledge tracing algorithms adapt to a particular topic, such as addition, subtraction, multiplication, and division.

\section{Smart Tutor App}

The frontend design is simplified to fit the scope of the dissertation. There is a lot of room to improve the design to make education more enticing, accessible, and possibly more effective (e.g., using games to teach mathematics in practise through game theory). In future work, this aspect could be emphasised, and research could be conducted into how effective using games is for teaching concepts in practice, such as maths or physics.

Another problem that arose from a tight scope is that the questions are limited to simple maths. Knowledge tracing can be applied to a wide range of topics, from simple maths to programming exercises and even modelling game skills, as long as the skill can be evaluated as correct or incorrect. While the API design could be expanded to cover more topics. This was beyond the scope of the dissertation.