\chapter{Methodology}
This section describes the approach used in this dissertation to achieve its aims and goals.

\section{Smart Tutor App}
The app comprises two parts: the front-end user interface (UI), which users interact with, and the backend server, which processes data and manages requests.

The backend is a Representational State Transfer (REST) API\@. It uses the pyKT~\autocite{pyktPaper} library to predict response correctness from historical data. This REST API architecture was chosen to enable communication between the Unity front end and the server. It also allows user data to be synthesised via a script.

The frontend was designed in Unity, a cross-platform game engine. This choice created a more user-friendly experience and enhanced the GUI\@. To communicate with the backend, it uses a coroutine-based network manager. An event bus publishes received responses as events. These events trigger UI changes and update the current question through subscribers. Unity's multi-platform support enables building the frontend for WebGL\@. This allows it to be hosted on the same server as the API\@. As a result, the frontend serves as the landing page when users visit the server's URL, seamlessly integrating with the deployment process. Building to WebGL also enables any device with a browser to access this website. Whether a device runs Android, iOS, macOS, Windows, or Linux, it can use the website if it has a working browser and a touchscreen or mouse.

Building on this integrated approach, the project is publicly accessible at \href{https://dissertation.jakubwj.com}{dissertation.jakubwj.com}. The server is hosted on a Raspberry Pi 5. A Cloudflare tunnel provides a secure connection between the server and the Internet. This tunnel routes traffic to an Nginx reverse proxy, which forwards requests to a Gunicorn systemctl service. Gunicorn acts as the Web Server Gateway Interface (WSGI) and enables communication between the server and the Flask backend application.

\subsection{Data Handling}
In addition to deployment considerations, the app uses an SQLite database. This database stores users, questions, responses, and other data, as shown in \figref{figure:database_diagram}. Student users can be synthesised, so they are captured as a source in the database. Currently, synthesisers are named after the large language model used to create the user data. Students without a source are real students.

For database management, questions are generated dynamically from a seed. Once a question is generated, the seed, question text, and correct answer get stored in the question database. The seed keeps track of the generator ID, generator version and question seed. For example the seed ``MUL:v1:96'' refers to the version 1 multiplication generator and the seed 96 corresponds to 6*9.

Each database question links to one or more knowledge components (skills). Currently, all generated questions are simple and fit into addition, subtraction, multiplication, or division.

The pyKT library trains models on CSV files in a pre-specified format. To enable training, the app includes two scripts: one that exports the database to a CSV file and another that preprocesses the data to the required format.

Users also store passwords and user IDs for authentication and authorisation. The passwords are hashed and salted using modern hashing algorithms (werkzeug.security). The session tokens are handled by the Python library Flask-JWT-Extended. The JWT tokens are signed using a secret generated with openssl. These tokens are sent in the API request headers. They enable prolonged sessions in which the user's identity is guaranteed. When users log out, their session is added to a JWT blacklist, preventing that token from being used to authenticate them. Once the token is placed into the blocklist, a new token must be retrieved (The user must log in again).

\section{User Data Synthesis}
Transitioning to data privacy considerations, the primary purpose of data synthesis is to avoid collecting sensitive data. Another implicit assumption is that data synthesis should be as close as possible to real user data. The user synthesis process is detailed in the flowchart in \figref{figure:synthesis_flowchart}.

To implement these synthesis aims, user synthesis is performed through a dedicated student synthesis subproject running in its own virtual environment. At first, an LLM is created using a custom model prompt. This prompt can be found in Listing~\ref{lst:sys-prompt}. The script iterates through all the selected models: LLama3.2:latest~\autocite{llama3}, Mistral:latest~\autocite{mistral7b}, Qwen2.5:latest~\autocite{qwen2.5}, Gemma3:1b~\autocite{gemma_2025}, and DeepSeek-R1:1.5b~\autocite{deepseek-r1}. The goal of the selection was to have a wide range of models to see whether any could predict user data significantly better than the others.

Once the synthesis subproject is configured, the script starts initialising students. Each user is created on a separate thread to increase performance. Each student has an internal state similar to Bayesian Knowledge Tracing~\autocite{bkt}. It randomly initialises the learn, slip, guess rate, and mastery for each skill $k$. The parameters are sampled to establish the student's initial state.

\subsection{Internal State Initialization}

After students are initialised, each user has a prior overall mastery $\pi_{L_0}$ and learning rate $\pi_T$. This was done to simulate relationships among the proficiencies of different skills. Stochastic noise $\epsilon_k$ is added to both the overall mastery and the learning rate per skill to differentiate mastery levels between skills. All parameters are clamped to realistic values.

The mastery threshold was set at 0.75. This accommodates students with high prior mastery while always leaving room for progression. The learning rate was reduced to 0.1 so students can make reasonable progress over about 80 questions. All parameters have minimum values because it is assumed that any parameter being 0 is unrealistic. The guess rate is lower than the slip rate. This is because real students are more likely to consciously guess than get a question they've mastered wrong. The guess rate is also higher because questions in the smart tutor app cannot be skipped. When a student does not know how to solve a problem, they are forced to guess.

\[\pi_{L_0} \sim \mathcal{U}(0, 0.75)\]
\[\pi_T \sim \mathcal{U}(0.01, 0.1)\]

For each skill, the following variables are assigned:

\[ \epsilon_k^{p(L_0)} \sim \mathcal{N}(0, 0.1), \quad {p(L_0)}_k = \mathrm{clip}(\pi_{L_0} + \epsilon_k^{p(L_0)}, 0.05, 0.75, 2)\]
\[ \epsilon_k^{p(T)} \sim \mathcal{N}(0, 0.025), \quad {p(T)}_k = \mathrm{clip}(\pi_T + \epsilon_k^{p(T)}, 0.01, 0.1, 2)\]
\[ {p(G)}_k \sim \mathrm{round}(\mathcal{U}(0.05, 0.2), 2)\]
\[ {p(S)}_k \sim \mathrm{round}(\mathcal{U}(0.01, 0.1), 2)\]

Where:
\begin{itemize}
    \item ${p(L_0)}_k$ is the initial mastery per skill, ${p(T)}_k$, ${p(G)}_k$, ${p(S)}_k$ are the initial learning, slip, and guess rates per skill, respectively
    \item $\mathcal{U}(a, b)$ is a random sample from a uniform distribution between $a$ and $b$
    \item $\mathcal{N}(a, b)$ is a random sample from a normal distribution between $a$ and $b$
    \item $\mathrm{round}(a, b)$ is a function that rounds $a$ to $b$ decimals
    \item $\mathrm{clip}(a, b, c, d)$ is a function that clamps $a$ between $b$ and $c$ and rounded to $d$ decimal digits
\end{itemize}

After initialising a student's state, the student is registered and logged in to the API\@. User credentials are created using Faker and saved to a text file after synthesis. The session token is stored in the student class for question retrieval and logging. On registration, a synthesiser is assigned to the student; the LLM name matches the synthesiser. If registration or login fails, the user is skipped.

Once the user is logged in, a question is retrieved. This question is used to construct the prompt for the large language model. The prompt includes all the skill states, a history of previously answered questions with their generated answers, and the current question. For each student, the number of questions is drawn uniformly at random from 5 to 100. This range is narrower than the pyKT library default (3--200) to reduce load and remove sequences that are too short.

\subsection{Updating the internal state}

To introduce more complexity into the data, all the skills and past answers were included in the LLM prompt. The prompt is sent to the LLM, and the model's response is validated. The validation only checks that the prompt is valid JSON with the required keys and types. It doesn't validate outlandish answers or response times. If the format is invalid, the large language model is prompted to generate the answer again. If the prompt is invalid 3 times in a row, the question is skipped.

If the response is successfully validated, the answer is logged to the server. The server returns whether the question is correct, and the student's internal state is updated accordingly. The mastery for question $t$ is updated by these two formulas:

When the answer is correct:
\[ P(L_{t+1}\ |\ \text{correct}) = \frac{{p(L_t)}_k \cdot (1 - {p(S)}_k)}{{p(L_t)}_k \cdot (1-{p(S)}_k) + (1-{p(L_t)}_k) \cdot {p(G)}_k} \]

When the answer is incorrect:
\[ P(L_{t+1}\ |\ \text{incorrect}) = \frac{{p(L_t)}_k \cdot {p(S)}_k}{{p(L_t)}_k \cdot {p(S)}_k + (1-{p(L_t)}_k) \cdot (1-{p(G)}_k)} \]

The question is added to the student's answer history. A new question is retrieved from the server, and a new cycle starts. The next LLM prompt includes the new question, the updated history, and the skill states. This cycle repeats until a desired number of questions is reached.

The student synthesis ends once all questions and answers are generated. Student credentials are then logged to a credentials file for possible later use.

\section{Model and Synthetic Data Comparison}
All the generated datasets (LLama3.2:latest, Qwen3.5:latest, Mistral, Qwen2.5:latest, and DeepSeek-R1) were trained using all the selected models (AKT, DKT, DTransformer, SAKT, and SimpleKT). To verify the correctness of the implementation, a truncated version of ASSISTments2009~\autocite{assist2009} was used. The SimpleKT model was used as the baseline, and ASSISTments2009 as the baseline dataset.

To ensure the same conditions across all knowledge tracing models and datasets, datasets are truncated to 100 users. The decision to limit the users to 100 was also influenced by performance considerations. The user state for all datasets is generated using the same function.

The ASSISTments2009 dataset is a collection of real user data; the questions and knowledge components are represented only by IDs. The real question and components aren't publicly available. This means that a direct comparison isn't possible for the trained data. Comparing these algorithms on real user data can show how well they perform.

The knowledge tracing models are compared based on their accuracy and AUC scores. These scores are the most common in knowledge tracing.~\autocites{akt, dkt, pyktPaper}. The AUC score is used in the knowledge tracing field because knowledge tracing is a binary prediction (the next answer is correct or incorrect), and AUC assesses the ranking quality of these predictions. AUC is also useful for ranking skills and questions in problem recommendation systems built on knowledge tracing. Accuracy is also commonly used because it's easier to interpret than AUC\@.

In addition to scoring the models on a test set, they are also scored on the last epoch's training and validation data. Checking training and validation data should reveal potential over-fitting.

\section{Ethical Considerations}

Knowledge Tracing often requires the collection of personal data, a challenge that is particularly acute in domains such as early mathematics, which primarily involve children. Collecting data from children raises significant ethical concerns because they are unable to give informed consent. To address these concerns, the dissertation avoids collecting data from children and instead simulates user interactions using LLMs.

The Smart Tutor system is designed to minimise privacy risks. While it distinguishes between users, it does not collect any personally identifiable information. Instead of using usernames, the system uses generated user IDs. The  passwords are hashed using modern hashing algorithms. The only data stored from user interactions are the question IDs, correctness Booleans, timestamps and response times, with no additional context or personal details. This approach ensures that user privacy is maintained throughout, while still enabling meaningful research into Knowledge Tracing methods.
