\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{geometry, float}
\usepackage[style=authoryear, backend=biber]{biblatex}
\usepackage{mhchem}
\usepackage{amsmath}
\usepackage{pgfgantt}

% tables
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ltablex}

\geometry{ a4paper, top=2cm, bottom=2cm, left=2cm, right=2cm, heightrounded}
\renewcommand{\baselinestretch}{1.15}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.8em}

\addbibresource{references.bib}

\newcommand{\figref}[1]{Figure~\ref{#1}}

\title{Dissertation Proposal: Comparing Knowledge Tracing Algorithms in Simple Maths Operations}
\author{Jakub Watson-Jones}
\date{10 February 2026}

\begin{document} \maketitle

\section{Context}

Maths education is a crucial part of everyone's lives. Its primary goal is to equip students with fluency in the fundamentals of mathematics, mathematical reasoning, and problem-solving, and to apply mathematical concepts in science and other fields~\autocite{Math_Curriculum_England}. Government reports suggest that the average class size is around 26.2 pupils~\autocite{Class_Size_England} and has an average pupil-teacher ratio of 18.0~\autocite{Pupil_Teacher_Ratio_England}. This high ratio inevitably limits the attention a teacher can provide to each student.

Learning is separated into two groups: procedural fluency and contextual understanding. In maths, there is a large emphasis on procedural fluency in the early stages of education. In the UK, pupils do mental arithmetic in Key Stage 1. This is why calculators are only introduced at the end of Key Stage 2~\autocite{Math_Curriculum_England}.

\subsection{Problem}

When teaching early arithmetic and multiplication tables in schools, the classical approach is to give pupils a long list of exercises. However, this approach is not optimal as a pupil can encounter problems that they have already mastered. Alternatively, they may encounter problems that they are unable to solve. A better approach is to keep the student in the Zone of Proximal development~\autocite{Zone_of_Proximal_Development}.

Keeping students in their Zone of Proximal Development is a known challenge, especially in large classes. A teacher can't adjust materials to each student individually. This is why teachers are forced to make trade-offs. To address this, I will use Knowledge Tracing. Knowledge tracing adjusts the materials based on a measured mastery of each topic. The mastery is the probability of the user answering correctly. This mastery is then used to estimate the Zone of Proximal Development.

Because Knowledge Tracing algorithms create an identifiable state of mastery, it comes with data privacy concerns. Specifically, the General Data Protection Regulation complicates the collection of training and testing data. Existing papers show the possibility of avoiding these concerns by synthesising user data. These papers use publicly available student data and extend it using different synthesis methods~\autocite{pagonis2024studentsynthesis}. The challenge will be adapting these methodologies to basic maths and response time.

\section{Aim}

The overall aim of my project is to create an educational artefact that enhances learning using Knowledge Tracing algorithms. I intend to train and test the effectiveness of the Knowledge Tracing algorithms using synthesised student data.

\begin{enumerate}
    \item I will create an artefact that has a user interface (UI) for human interactions and an application programming interface (API) for synthetic users. Both the UI and API should accept a score and user response time, preprocess them and send them to the Knowledge Tracing algorithms.
    \item The artefact should be able to handle multiple Knowledge Tracing algorithms. This should enable their comparison.
    \item I will implement a user synthesis system and expand it to generate user response times for answers. A working generator for synthetic user data should be able to create arbitrarily many answer scores and response times.
    \item The artefact will be evaluated on the synthetic student data. The knowledge tracing algorithm comparison will be based on the synthetic user data. Using synthetic user data as opposed to real user data is controlled and repeatable.
\end{enumerate}

\subsection{Key Background Sources}


\begin{tabularx}{\textwidth}
    {
        >{\raggedright\arraybackslash}X
        >{\raggedright\arraybackslash}X
        >{\raggedright\arraybackslash}X
    }
    \toprule
    Paper                                                                                                                                                                                                                                                                                                       & Summary & Relationship with project \\
    \midrule\endhead{}
    Analysis of Knowledge Tracing performance on synthesised student data~\autocite{pagonis2024studentsynthesis}                                                                                                                                                                                               &
    This paper compares different methods of expanding student data. The comparison is based on the performances of Bayes and Deep Knowledge Tracing.                                                                                                                                                           &
    This paper will serve as a basis for creating synthetic data to test and train the Knowledge Tracing algorithms.                                                                                                                                                                                                                                  \\

    \midrule
    A Survey of Knowledge Tracing: Models, Variants, and Applications~\autocite{Shen_2024}                                                                                                                                                                                                                     &
    This paper provides an overview of Knowledge Tracing algorithms, and provides an open source algorithm library called EduKTM, and KT-datasets EduData                                                                                                                                                       &
    I intend to use the knowledge tracing algorithm library to test out multiple algorithms in my project.                                                                                                                                                                                                                                            \\

    \midrule
    Learn with M.E.—Let Us Boost Personalized Learning in K-12 Math Education!~\autocite{educsci14070773}                                                                                                                                                                                                      &
    This article describes an existing artefact concerned with teaching basic mathematical operations. The artefact uses rule-based difficulty adjustment and focuses on analysing mistakes by comparing them to a bank of 429 error patterns. The paper shows improvement in learning in real-world scenarios. &
    This paper serves as a baseline for my dissertation.  Their artefact uses a rule-based system to adjust difficulty, which will be a good comparison to the Knowledge Tracing algorithms.                                                                                                                                                          \\

    \midrule
    Knowledge tracing: Modelling the acquisition of procedural knowledge~\autocite{CorbettAlbertT.1995KtMt}                                                                                                                                                                                                    &
    This paper introduces a widespread algorithm called Bayes Knowledge Tracing (BKT).                                                                                                                                                                                                                          &
    I'll implement this model, as it's one of the most influential algorithms.                                                                                                                                                                                                                                                                        \\

    \midrule
    Deep Knowledge Tracing~\autocite{piech2015deepknowledgetracing}                                                                                                                                                                                                                                            &
    This influential paper describes a modern adaptation of Knowledge Tracing using Machine Learning.                                                                                                                                                                                                           &
    I intend to implement and expand this algorithm to use response time                                                                                                                                                                                                                                                                              \\

    \bottomrule
\end{tabularx}

\section{Planning}
\subsection{Gantt Chart}

\ganttset{
    calendar week text={
            \number\startday/\number\startmonth
        },
}
\begin{ganttchart}[
        vgrid={*6{thin, dotted}, *1{thick, dashed}},
        expand chart=\textwidth,
        y unit title=7.5mm,
        title height=1,
        y unit chart=5mm,
        time slot format=little-endian,
        bar label font=\small\upshape\mdseries,
        milestone label font=\small\upshape\mdseries,
        milestone left shift=-0.6,
        milestone right shift=-0.4,
        milestone height=0.3,
        bar/.style={
                rounded corners=2pt,
                draw=black!90,
                top color=white,
                bottom color=red!75,
            },
    ]{16/2/2026}{10/5/2026}
    \gantttitlecalendar{month, week} \\
    \ganttbar[bar/.append style={bottom color=blue!75}]{Writing}{23/02/2026}{7/05/2026} \\
    \ganttbar{Create artefact skeleton}{20/2/2026}{3/3/2026} \\
    \ganttmilestone{Mid-project check-in}{4/3/2026} \\
    \ganttbar{Implement user synthesis}{4/3/2026}{11/3/2026} \\
    \ganttbar{Implement Knowledge Tracing}{12/3/2026}{19/3/2026} \\
    \ganttbar{Create Poster}{16/3/2026}{18/3/2026} \\
    \ganttmilestone{Poster due}{20/3/2026} \\
    \ganttbar[bar/.append style={bottom color=green!75}]{Debug \& Catch-up}{20/3/2026}{29/3/2026} \\
    \ganttbar{Implement response time}{30/3/2026}{12/4/2026} \\
    \ganttbar{Testing \& Evaluation}{13/4/2026}{19/4/2026} \\
    \ganttbar{Demo preparation}{20/4/2026}{26/4/2026} \\
    \ganttbar[bar/.append style={bottom color=green!75}]{Easter Break}{21/3/2026}{20/4/2026} \\
    \ganttmilestone{Dissertation due}{06/5/2026} \\
    \ganttmilestone{Demonstration Fair}{07/5/2026}
\end{ganttchart}

\subsection{Work Plan Explanation}

For this project, I plan to use an iterative approach. At first, I will start by fleshing out a rough skeleton of the artefact. This initial artefact should work from start to finish. It should have an initial UI and a Knowledge Tracing algorithm. I plan to start writing my dissertation after I begin coding to have something to write about. Around the mid-project check-in, I should have a working prototype. At this point, I should be preparing to implement user synthesis. Additionally, I hope to have written an introduction. I should also have begun working on the background research and literature review.

After the mid-project check-in, I will implement user synthesis and multiple knowledge tracing algorithms. I plan to implement the Bayes, Deep, and other Knowledge Tracing algorithms. It's important to finish a version of the user synthesis and knowledge tracing algorithms, so I have some data for the poster due on the $20^{th}$ of March.

During Easter, I intend to take a short break and do some debugging and testing. Once I come back, I intend to implement response time into both the knowledge tracing and the synthetic users.

At the end of my project, I will test and evaluate both the knowledge tracing and the user synthesis. Once I'm done testing, I will have time to finish writing my dissertation and polish the artefact. I will also have time to prepare for presenting at the Demonstration Fair. If I have some spare time, I would like to gamify the UI\@.

\subsection{Risk}
\begin{itemize}
    \item One risk I've identified is the implementation of response time. I recognise that this might be a challenge and will not be within the scope of this project. To ensure I don't fall behind, I intend to implement response time as the last part of the project. If the response times' implementation isn't complete within its allocated time, I will drop it from the project and base the results on the correctness of the responses.
    \item Another risk is that I fall ill or am unable to work on the project. To address this risk, I will develop the artefact iteratively, so that at any point, I will have a working project ready to present. I've also left two windows, where I can catch up if necessary. First at the beginning of Easter and the other towards the end of the project.
\end{itemize}

\newpage
\printbibliography{}
\end{document}